---
title: 'Screening model for LUNG CANCER'
author: "Yi_Ping"
output: html_document
---
### import library
```{r}
### import the library 
library(readxl)
library(dplyr)
library(factoextra) 
library(ggplot2)
library(pROC)
library(stats)
library(caret)
library(pheatmap)
library(clusterProfiler)
library(ggstatsplot)
### import the function
source('./used_function.R')

```

### import dataset
```{r warning=FALSE}
### tss matrix
BJYYBIElung_tssmatrix <- read.csv("~/Desktop/BDF/data/BJYYBIElung_tssmatrix.csv")
dim(BJYYBIElung_tssmatrix) # 93701   224
### change '.' into '-'
colnames(BJYYBIElung_tssmatrix)[3:ncol(BJYYBIElung_tssmatrix)] <- 
gsub('[.]','-',colnames(BJYYBIElung_tssmatrix)[3:ncol(BJYYBIElung_tssmatrix)])
### clinic data
lung_clinic <- read_excel("~/Desktop/BDF/data/lung_clinic_used.xlsx")
dim(lung_clinic) # 96  8

unique(colnames(BJYYBIElung_tssmatrix))
colnames(BJYYBIElung_tssmatrix)# 1:2, 3:124 ,125:224
```

### clinical information for this dataset
```{r}
lung_clinic_used <- lung_clinic[1:96,]
lung_clinic_used$Nmeta <- as.numeric(lung_clinic_used$Nmeta)
table(lung_clinic_used$Nmeta)
#0  1  2  3  4  5  7 
#57 21  8  4  2  3  1 
table(lung_clinic_used$Nnodes)
# 0  1  2  3 
#45  2 18 31 
table(lung_clinic_used$stage_less)
# I II III IV
# 32 3 22 39
table(lung_clinic_used$GGOorsolid)
# ggo           NA solid nodule 
# 20           61           15 
```

### pre_treat 
```{r}
### 96 samples selected
lung_tss <- BJYYBIElung_tssmatrix[,lung_clinic_used$sample]
lung_tss <- cbind(lung_tss,BJYYBIElung_tssmatrix[,125:224])
dim(lung_tss) # 93701   196
### pre_treat the tss matrix
lung_tss_pred <- pre_treat_rm0pNA(lung_tss)
dim(lung_tss_pred)# 196 90808

```

###########################################################################
########################## for CA nonCA model #############################
###########################################################################
### 70% training & 30% validation 
### normalize for sites & save the normalization of training for further use
```{r}
model_clinic <- data.frame(sample = c(lung_clinic_used$sample,colnames(BJYYBIElung_tssmatrix[,125:224])))
model_clinic$group = 'CA'
model_clinic$group[97:nrow(model_clinic)] = 'nonCA'
### 70% for training 30% for testing
  clinic = model_clinic
  matrix = lung_tss_pred
  set.seed(123)
  inTrain=createDataPartition(y=clinic$group,p=0.70,list=FALSE)
  training=matrix[inTrain,]
  testing=matrix[-inTrain,]
  ### save the group of training && testing 
  trainLabels <- clinic[inTrain,'group']
  testLabels <- clinic[-inTrain,'group']
  test_clinic_matrix<- clinic[-inTrain,]
  dim(training) #   147 90808
  dim(testing)  #  49 90808
  
  training_range_sample <- range_cal(matrix = training,roworcol = 1)
  range(training_range_sample$median) #  0.596288 0.958518
  
  training_range_site <- range_cal(matrix = training,roworcol = 2)
  range(training_range_site$median) # 0.000 170.935
  
  ### do the normalize for the sites -- use the z-score && median & MAD and ignore the outliers
  training_range_site <- training_range_site[which(training_range_site$mad != 0),]
  site_used <- rownames(training_range_site[which(training_range_site$mad != 0),])
  
  training_rmmad <- training[,site_used]
  
  training_zscore <- data.frame(apply(data.frame(training_rmmad), 2, custom_z_score))
  training_zscore_range <- range_cal(matrix = training_zscore,roworcol = 2)
  training_zscore_rangesamples <- range_cal(matrix = training_zscore,roworcol = 1)
  range(training_zscore_range$median) # -3.957338e-16  7.216450e-16
  dim(training_zscore) #   147 90794
  
  testing_zscore <- newmatrix_z_norm(new_matrix=testing,training_range_site)
  dim(testing_zscore) #  53 90599
  testing_zscore_range <- range_cal(matrix = testing_zscore,roworcol = 2)
  range(testing_zscore_range$median) #  -0.9618575 25.8452444
```

### Dimensionality reduction for early screening model
### STEP1: use the training set to calculate the p-value of sites, only p<0.01 will be remained  
### STEP2: 1. calculate the rank score of every patient for every site 2.to calculate the mean of mean rank score of several groups (healthy,stageI,stageII,stageIII,stageIV) 3. only choose the sites with the same trend from early stages to advanced stages 4. sort by gap between healthy and stageI && choose top 200 features

### rank values sum (use clinic_rank && train_topbins)
```{r}
### STEP1: p<0.01
group_train = data.frame(disease = clinic[inTrain,'group'])
dim(group_train)
tss_z_pCA <- p_cal(trainTransformed = training_zscore,
                     trainLabels = group_train$disease,
                     trainLabels_sig = 'CA',paired = FALSE)  
tss_z_pCAcut <- tss_z_pCA[which(tss_z_pCA[,1] < as.numeric(0.01) ),]
dim(tss_z_pCAcut)  #  19650     2
train_topbins<-as.data.frame(training_zscore)[,tss_z_pCAcut$V2]
### STEP2:
### rank for every feature
clinic_rank <- lung_clinic_used[,c('sample','stage_less')]
### rank matrix for train_topbins
train_topbins_rank <- data.frame(apply(train_topbins,2,function(x) return(rank(x))))
train_topbins_rank$sample = rownames(train_topbins_rank)
### sum of several groups
train_topbins_rank <- left_join(train_topbins_rank,clinic_rank,by='sample')
colnames(train_topbins_rank)[ncol(train_topbins_rank)]

train_topbins_rank[is.na(train_topbins_rank$stage_less),'stage_less'] = 'healthy'
Nsites = ncol(train_topbins_rank)-2
### calculate the mean
group_mean <- function(value_col){
  matrix_col <- data.frame(stage_less = train_topbins_rank$stage_less,
                           value = value_col)
  result_matrix <- data.frame(aggregate(value ~ stage_less,data = matrix_col, FUN = mean))
  return(result_matrix$value)
}

result_matrix_mean <- data.frame(site=colnames(train_topbins_rank)[1:Nsites])
result_matrix_mean$healthy = 1
result_matrix_mean$I = 1
result_matrix_mean$II = 1
result_matrix_mean$III = 1
result_matrix_mean$IV = 1

for (i in 1:Nsites) {
  result_matrix_mean[i,2:6] <- group_mean(train_topbins_rank[,i])
}

### rm stage II because just includes 2 samples 
rownames(result_matrix_mean) <- result_matrix_mean$site
result_matrix_mean2 <- result_matrix_mean[,c(2,3,5,6)]
### only the up or down trend of (healthy,I,III,IV)
up_sort <- function(one_row){
  one_row =  as.numeric(one_row)
  return(all(diff(one_row) > 0))
}

down_sort <- function(one_row){
  one_row =  as.numeric(one_row)
  return(all(diff(one_row) < 0))
}
up_sort(result_matrix_mean2[3,])
result_matrix_mean2$up <- apply(result_matrix_mean2,1,up_sort)
result_matrix_mean2$down <- apply(result_matrix_mean2,1,down_sort)
### up && down
result_matrix_mean3 <- result_matrix_mean2[which(result_matrix_mean2$up == TRUE | result_matrix_mean2$down == TRUE),] # 2402
result_matrix_mean3 <- result_matrix_mean2[which(result_matrix_mean2$up == TRUE),]#2226

result_matrix_mean4 <- result_matrix_mean2[which(result_matrix_mean2$up == TRUE | result_matrix_mean2$down == TRUE),]

result_matrix_mean3$gap = result_matrix_mean3[,2]-result_matrix_mean3[,1]
result_matrix_mean3$gap2 = result_matrix_mean3[,4]-result_matrix_mean3[,1]
result_matrix_mean3$gap3 = result_matrix_mean3[,4]-result_matrix_mean3[,2]

### sort by gap && choose top 200
result_matrix_mean3 <- result_matrix_mean3[order(-result_matrix_mean3$gap),]
result_matrix_mean3_cut <- result_matrix_mean3[1:as.numeric(200),]

rank_sites <- rownames(result_matrix_mean3_cut)
length(rank_sites) # 200
```
### use the 200 fetures to build the model
```{r}
### build the model based on features:rank_sites
  train_topbins<-as.data.frame(training_zscore)[,rank_sites]
  dim(train_topbins) #  158 300
  test_topbins<-as.data.frame(testing_zscore)[,rank_sites]
  dim(test_topbins) # 52 300
  
  train_topbins$group <- factor(trainLabels,levels = c('CA','nonCA'))
  test_topbins$group <- factor(testLabels,levels = c('CA','nonCA'))
  used_100f <- rbind(train_topbins,test_topbins)
  
  ### build model based on rf
  set.seed(123) 
  ctrl = trainControl(method = "repeatedcv", number = 10, repeats = 3,
                      allowParallel = TRUE,classProbs = TRUE,savePredictions = TRUE,search='grid',summaryFunction = twoClassSummary)
  set.seed(123) 
  ctrl$sampling <- "up"
  tunegrid <- expand.grid(mtry = c(1))
  set.seed(123)
  ### 0.9027833   0.9027
  rf_default1 <- train(group~., 
                       data=train_topbins, 
                       method='rf', 
                       #metric='Accuracy',  #Metric compare model is Accuracy
                       metric = "ROC",
                       tuneGrid=tunegrid,
                       ntree=c(1500),
                       trControl=ctrl)
  rf_default1
  modeltss_CA <- rf_default1
  rf_default1 <- modeltss_CA
  CAnonCAp=tss_z_pCAcut #19650

  train_topbins<-as.data.frame(training_zscore)[,rank_sites]
  dim(train_topbins) #  158 300
  test_topbins<-as.data.frame(testing_zscore)[,rank_sites]
  dim(test_topbins) # 52 300
  
  ### prediction for the 30% validation data
  CAnontesting_zscore <- test_topbins 
  test_topbins <- CAnontesting_zscore 
  pred <- predict(rf_default1,newdata = test_topbins,type = "prob")
  pred$sample = rownames(pred)
  pred <- left_join(pred,model_clinic,by='sample')
  pred$res = 1
  pred[which(pred$CA<0.4),'res'] = 0 
  
  pred$true = 1
  pred[which(pred$group != 'CA'),'true'] = 0 
  
  pred$equ = 1
  pred[which(pred$res != pred$true),'equ'] = 0 
  roc_data <- roc(pred$true, pred[,1],type="prob")
  print(roc_data)
  pred <- left_join(pred,lung_clinic_used,'sample')
  
```
### save data of CA model
```{r}
### save data of CA model
CAnonCAp=tss_z_pCAcut #19650
modeltss_CA200 = modeltss_CA 
CAtrain = training_zscore
CAtest = testing_zscore
load('/Users/pingyi/Desktop/BDF/script/out_data/CAnonCA200f')
save(training_range_site,testing,CAnonCAp,result_matrix_mean4, modeltss_CA200,CAtrain,CAtest,
     file = '/Users/pingyi/Desktop/BDF/script/out_data/CAnonCA200f')
#save(used_100f,modeltss_CA,file = '/Users/pingyi/Desktop/BDF/script/out_data/CAnonCA100f')

```


